---
title: "Prediction Analysis"
date: "2024-04-30"
output: 
  html_document: 
    code_folding: show
    highlight: haddock
    number_sections: yes
    theme: lumen
    toc: yes
    toc_depth: 4
    toc_float: yes
  pdf_document: 
    number_sections: yes
    toc: yes
    toc_depth: '4'
  word_document: 
    toc: yes
    toc_depth: '4'
urlcolor: blue
---

```{r setup, include=FALSE}
# Remove all existing objects from environment
rm(list = ls())

# Load libraries
library("tidyverse") # open source packages for tidy data
library("ggsci") # color palettes from scientific journals
library("sandwich") # calculating HC standard errors
library("car")
library("stargazer")
library("skimr") # generate easy data summaries
library("broom")
library("pander")

# Set dplyr::select function as the default function
select <- dplyr::select

# Set plot themes
theme_set(theme_classic())
options(
  ggplot2.discrete.colour = ggsci::scale_colour_npg,
  ggplot2.discrete.fill = ggsci::scale_fill_npg
)

# Set option so that scientific notation isn't displayed
options(scipen = 999)

```

# INGEST DATA

```{r ingest_and_clean_data, include=FALSE}

# Ingest Data ------------------------------------------------------------------

# Read .csv file
vaccination_raw <- read_csv("vaccination-30-days.csv")

# Create Day, Time, and Condition Factors --------------------------------------

# --- Create labels for factors

# Define the weekdays in order
weekdays <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")

# Define times in 30-minute intervals
times <- sprintf("%02d:%02d", rep(0:23, each = 2), rep(c(0, 30), times = 24))
times_starting_noon <- c(times[25:48], times[1:24])

# Create all combinations of weekdays and times
combos <- expand.grid(Day = weekdays, Time = times)

# Order by weekday first, then by time
combos <- combos[order(combos$Day, combos$Time), ]

# Combine Day and Time into a single string
time_intervals <- with(combos, paste(Day, Time, sep = "_"))

# Define factors for days
days <- c("day_1", "day_2", "day_3")

# Change variables into factors for each file
# Add missing indicators for medicare and medicaid
vaccination_all_fields <- vaccination_raw |>
  mutate(
    last_vax_time_30_min = substr(last_vax_time_30_min, 1, 5),
    last_vax_time_30_min = factor(last_vax_time_30_min, levels = times),
    last_vax_dow_time_30_min = factor(last_vax_dow_time_30_min, levels = time_intervals),
    last_vax_dow_30_min = factor(last_vax_dow_30_min, levels = weekdays),
    insurance_commercial = ifelse(insurance_medicaid == 0 & insurance_medicare == 0 & insurance_other == 0, 1, 0),
    insurance_missing = ifelse(is.na(insurance_medicaid) | is.na(insurance_medicare) | is.na(insurance_other), 1, 0),
    insurance_medicaid = ifelse(insurance_missing == 1, 0, insurance_medicaid),
    insurance_medicare = ifelse(insurance_missing == 1, 0, insurance_medicare),
    condition = factor(condition, levels = c("control", "popular", "personal")),
    control = ifelse(condition == "control", 1, 0),
    day_3 = ifelse(day_1 == 0 & day_2 == 0, 1, 0),
    insurance = factor(
      case_when(
        insurance_commercial == 1 ~ "commercial", 
        insurance_medicaid == 1 ~ "medicaid",
        insurance_medicare ==1 ~ "medicare",
        insurance_other == 1 ~ "other",
        insurance_missing == 1 ~ "missing"
        )
      ),
    day_of_text = factor(
      case_when(
        day_1 == 1 ~ "day_1",
        day_2 == 1 ~ "day_2",
        day_3 == 1 ~ "day_3"),
      levels = days),
    timezone = factor(timezone)
  ) |>
  relocate(insurance, insurance_commercial, .before = insurance_medicaid) |>
  relocate(insurance_missing, .after = insurance_other) |>
  relocate(control, .before = popular) |>
  relocate(day_3, .after = day_2) |>
  relocate(day_of_text, .before = day_1)

# Drop rows with missing data (57 obserations where there is no last vax time)
rows_with_missing <- which(apply(vaccination_all_fields, 1, function(x) any(is.na(x))))
vaccination_all_fields <- vaccination_all_fields[-rows_with_missing,]

# Generate subset of vaccination data with indicator variables only
vaccination_ind <- select(
  vaccination_all_fields, 
  -condition, 
  -insurance,
  -day_of_text,
  -last_vax_dow_time_30_min, 
  -last_vax_time_30_min, 
  -last_vax_dow_30_min, 
  -timezone
  )

# Generate subset of vaccination data with categorical variables
vaccination <- vaccination_all_fields |>
  select( 
    -control, -popular, -personal,
    -insurance_commercial, -insurance_medicare, -insurance_medicaid, -insurance_other, -insurance_missing,
    -day_1, -day_2, -day_3,
    -last_vax_dow_time_30_min,
  ) |>
  mutate(
    last_vax_time_30_min = factor(last_vax_time_30_min, levels = times_starting_noon)
    )

```

# Exploratory Data Analysis

## Spearman's Correlation Table

```{r bc_30_day, include=FALSE}

# Generate a correlation matrix
cor_matrix <- cor(vaccination_ind, method = "spearman")

# Reshape the correlation matrix
cor_reshape <- melt(cor_matrix)

# Create a color palette with red for negative correlations and blue for positive correlations
col_palette <- colorRampPalette(c("#3C5488FF", "white", "#E64B35FF"))(n = 100)

# Open a PNG graphics device
png("spearman-correlation-plot.png", width = 800, height = 800)

# Use corrplot to plot the correlation matrix and color code it
corrplot(cor_matrix, type = "upper", tl.col = "black", col = col_palette,
         addCoef.col = "black", number.cex = 0.8, mar = c(0,0,1,0))

dev.off()

```

## Generate Barplots for specific variables

```{r, include=FALSE}

# Boxplot on Prev Flu Vax Count
ggplot(vaccination_all_fields, aes(x = factor(flu_vax_30_days), y = prev_flu_vax_count)) +
  geom_boxplot(fill = "#3B4992FF", outlier.shape = NA) + 
  labs(x = "Vaccination after 30 Days", y = "Number of Past Flu Shots in 8 Years")

ggsave("past-flu-shots-boxplot.png", width = 8, height = 4.5, dpi = 500)

# Boxplot on Age
ggplot(vaccination_all_fields, aes(x = factor(flu_vax_30_days), y = age)) +
  geom_boxplot(fill = "#3B4992FF", outlier.shape = NA) + 
  labs(x = "Vaccination after 30 Days", y = "Age")

ggsave("age-boxplot.png", width = 8, height = 4.5, dpi = 500)

```

## Mosaic plot

```{r, include=FALSE}

library(vcd)

# Generate dataframe for mosaic plots
vaccination_mosaic <- vaccination_all_fields |>
  select(flu_vax_30_days, insurance_medicare, flu_vax_prev_season) |>
  mutate(
    flu_vax_30_days = ifelse(flu_vax_30_days == 1, "Yes", "No"),
    flu_vax_prev_season = ifelse(flu_vax_prev_season == 1, "Yes", "No"),
    insurance_medicare = ifelse(insurance_medicare == 1, "Yes", "No"),
    flu_vax_30_days = factor(flu_vax_30_days),
    flu_vax_prev_season = factor(flu_vax_prev_season),
    insurance_medicare = factor(insurance_medicare)
  )

# Create a contingency table
tab <- table(vaccination_mosaic$flu_vax_30_days, vaccination_mosaic$flu_vax_prev_season)

# Save PNG 1 
png("mosaic-plot-vaccination-prev-flu-vax.png", width = 1200, height = 800, res = 200)

# Create mosaic plot 1
vcd::mosaic(flu_vax_30_days ~ flu_vax_prev_season, 
            data = vaccination_mosaic,
            direction = c("v", "h")
            )

dev.off()

# Save PNG 2
png("mosaic-plot-vaccination-medicare.png", width = 1200, height = 800, res = 200)

# Create mosaic plot 2
vcd::mosaic(flu_vax_30_days ~ insurance_medicare, 
            data = vaccination_mosaic,
            direction = c("v", "h")
            )

dev.off()


```


## Heatmap of past vaccination times

```{r}

times_open <- sprintf("%02d:%02d", rep(8:21, each = 2), rep(c(0, 30), times = 14))

df_heatmap_all <- vaccination_all_fields |>
  filter(flu_vax_30_days == 1) |>
  group_by(last_vax_dow_30_min, last_vax_time_30_min) |>
  summarise(count = n()) |>
  ungroup() |>
  mutate(total = sum(count), percent = count / total) |>
  filter(last_vax_time_30_min %in% times_open)

# Create the heatmap
ggplot(df_heatmap_all, aes(x = last_vax_dow_30_min, y = last_vax_time_30_min, fill = percent)) +
  geom_tile() +  # Create the tiles
  scale_fill_gradient(low = "white", high = "#3B4992FF", guide = guide_colourbar(ticks = T)) +  # Set the color gradient
  labs(fill = "% of Total Appointments", x = "Day of Week", y = "Time of Day") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Improve x-axis labels readability
  scale_y_discrete(limits = rev(times_open))

ggsave("last-vaccination-heatmap.png", width = 8, height = 4.5, dpi = 500)

```

# Divide Dataset into 60% Training Data, 20% Testing Data, and 20% Final Validation 

```{r}

# Set partitions
n_total <- nrow(vaccination)
n_train <- round(n_total * 0.6)
n_test <- round(n_total * 0.2)
n_train_and_test <- n_train + n_test
n_validation <- n_total - n_train_and_test

# Create separate datasets
set.seed(488)
data_train <- vaccination[sample(n_total, n_train, replace = FALSE), ]
data_not_train <- anti_join(vaccination, data_train)

set.seed(2002)
data_test <- data_not_train[sample(nrow(data_not_train), n_test, replace = FALSE), ]
data_validate <- anti_join(data_not_train, data_test)

data_train_and_test <- rbind(data_train, data_test)

library(glmnet)
library(caret)

# Also extract everything in matrix form
data_train_X <- model.matrix(~., data = data_train[,-1])
data_train_Y <- data_train$flu_vax_30_days
data_test_YX <- model.matrix(~., data = data_test)[,-1]

# Also extract everything with Y as factor for random forest ranger package
data_train_rf <- data_train
data_train_rf$flu_vax_30_days <- factor(data_train_rf$flu_vax_30_days)

```

# Predictive Classifiers

## OLS

```{r ols, echo=FALSE}

library(pROC)

# Run OLS
fit_ols <- lm(flu_vax_30_days ~ ., data = data_train)

# Output summary
summary(fit_ols)

# Obtain the predicted probabilities from the model
predicted_prob <- predict(fit_ols, data_test, type = "response")
predicted_prob <- ifelse(predicted_prob < 0, 0, predicted_prob)

# Plot ROC curve and calculate AUC
roc_ols <- roc(data_test$flu_vax_30_days, predicted_prob)
plot(roc_ols)

# Calculate the auc_value
auc_value <- auc(roc_ols)

# Establish a threshold at the optimal point on the ROC curve
threshold <- 0.5

# Use threshold to classify predictions
predicted_vax <- ifelse(predicted_prob > threshold, 1, 0)

# Create the confusion table
confusion_table <- table(Predicted = predicted_vax, Actual = data_test$flu_vax_30_days)
confusion_table

dput(confusion_table)

# Calculate the misclassification error
misclassification_error <- mean(predicted_vax != data_test$flu_vax_30_days)
print(paste("Misclassification Error:", misclassification_error))
```

## Logistic Regression

```{r}

# Run Logit
fit_logit <- glm(flu_vax_30_days ~ ., family = binomial(link = "logit"), 
                 data = data_train)

# Output summary
summary(fit_logit)

# Obtain the predicted probabilities from the model
predicted_prob <- predict(fit_logit, data_test, type = "response")
predicted_prob <- ifelse(predicted_prob < 0, 0, predicted_prob)

# Plot ROC curve and calculate AUC
roc_logit <- roc(data_test$flu_vax_30_days, predicted_prob)
plot(roc_logit)

# Calculate the auc_value
auc_value <- auc(roc_logit)

# Establish a threshold at the optimal point on the ROC curve
threshold <- 0.5

# Use threshold to classify predictions
predicted_vax <- ifelse(predicted_prob > threshold, 1, 0)

# Create the confusion table
confusion_table <- table(Predicted = predicted_vax, Actual = data_test$flu_vax_30_days)
confusion_table

dput(confusion_table)

# Calculate the misclassification error
misclassification_error <- mean(predicted_vax != data_test$flu_vax_30_days)
print(paste("Misclassification Error:", misclassification_error))
```

## Relaxed Lasso w/ OLS and Logistic Regression

```{r relaxed_lasso, echo=FALSE}

# Set the seed
set.seed(153)

# Fit the Lasso model
fit_lasso <- cv.glmnet(data_train_X, data_train_Y, alpha = 1, nfolds = 15, intercept = F)

# Plot the lasso fit
plot(fit_lasso)

# Get coefficients for lambda.1se
coef_1se <- coef(fit_lasso, s="lambda.1se")  
coef_1se <- coef_1se[which(coef_1se !=0),] 
var_1se <- rownames(as.matrix(coef_1se))[-1] # output the names  dim(as.matrix(coef.min))

# Prepare the dataset for a relaxed lasso with logistic regression
data_train_subset <- cbind(data_train_Y, data_train_X[,var_1se]) |>
  as_tibble() |>
  rename(flu_vax_30_days = data_train_Y)

# LOGIT RELAXED LASSO ----------------------------------------------------------

# Run Logit
fit_relaxed_lasso_logit <- glm(flu_vax_30_days ~ ., family = binomial(link = "logit"), 
                 data = data_train_subset)

# Output summary
summary(fit_relaxed_lasso_logit)

# Obtain the predicted probabilities from the model
predicted_prob <- predict(fit_relaxed_lasso_logit, as_tibble(data_test_YX), type = "response")

# Plot ROC curve and calculate AUC
roc_relaxed_lasso_logit <- roc(data_test$flu_vax_30_days, predicted_prob)
plot(roc_relaxed_lasso_logit)

# Calculate the auc_value
auc_value <- auc(roc_relaxed_lasso_logit)

# Establish a threshold at the optimal point on the ROC curve
threshold <- 0.5

# Use threshold to classify predictions
predicted_vax <- ifelse(predicted_prob > threshold, 1, 0)

# Create the confusion table
confusion_table <- table(Predicted = predicted_vax, Actual = data_test$flu_vax_30_days)
confusion_table

dput(confusion_table)

# Calculate the misclassification error
misclassification_error <- mean(predicted_vax != data_test$flu_vax_30_days)
print(paste("Misclassification Error:", misclassification_error))

# OLS RELAXED LASSO ---------------------------------


# Run Logit
fit_relaxed_lasso_ols <- lm(flu_vax_30_days ~ ., data = data_train_subset)

# Output summary
summary(fit_relaxed_lasso_ols)

# Obtain the predicted probabilities from the model
predicted_prob <- predict(fit_relaxed_lasso_ols, as_tibble(data_test_YX), type = "response")
predicted_prob <- ifelse(predicted_prob < 0, 0, predicted_prob)

# Plot ROC curve and calculate AUC
roc_relaxed_lasso_ols <- roc(data_test$flu_vax_30_days, predicted_prob)
plot(roc_relaxed_lasso_ols)

# Calculate the auc_value
auc_value <- auc(roc_relaxed_lasso_ols)

# Establish a threshold at the optimal point on the ROC curve
threshold <- 0.5

# Use threshold to classify predictions
predicted_vax <- ifelse(predicted_prob > threshold, 1, 0)

# Create the confusion table
confusion_table <- table(Predicted = predicted_vax, Actual = data_test$flu_vax_30_days)
confusion_table

dput(confusion_table)

# Calculate the misclassification error
misclassification_error <- mean(predicted_vax != data_test$flu_vax_30_days)
print(paste("Misclassification Error:", misclassification_error))

```

## Random Forest

```{r}

library(ranger)
set.seed(1111)

# Run Random Forest Algorithm
fit_rf <- ranger(flu_vax_30_days~., data_train_rf, mtry = 4, 
                    num.trees = 500, splitrule = "gini",
                    importance = "impurity", probability = T)

# Obtain the predicted probabilities from the model
predicted_prob <- predict(fit_rf, data_test, type = "response")$predictions[,2]

# Plot ROC curve and calculate AUC
roc_rf <- roc(data_test$flu_vax_30_days, predicted_prob)
plot(roc_rf)

# Calculate the auc_value
auc_value <- auc(roc_rf)


# Establish a threshold at the optimal point on the ROC curve
threshold <- 0.5

# Use threshold to classify predictions
predicted_vax <- ifelse(predicted_prob > threshold, 1, 0)

# Create the confusion table
confusion_table <- table(Predicted = predicted_vax, Actual = data_test$flu_vax_30_days)
confusion_table

dput(confusion_table)

# Calculate the misclassification error
misclassification_error <- mean(predicted_vax != data_test$flu_vax_30_days)
print(paste("Misclassification Error:", misclassification_error))

```

## Neural Network

```{r}

library(caret)
library(neuralnet)
library(nnet)
set.seed(221)

data_train$flu_vax_30_days <- factor(data_train$flu_vax_30_days)
data_test$flu_vax_30_days <- factor(data_test$flu_vax_30_days)

fit_nn <- nnet(flu_vax_30_days ~ ., data = data_train, size = 10, maxit = 100)

# Make predictions on test data
predicted_prob <- predict(fit_nn, data_test, type = "raw")

# Plot ROC curve and calculate AUC
roc_nn <- roc(data_test$flu_vax_30_days, predicted_prob)
plot(roc_nn)

# Calculate the auc_value
auc_value <- auc(roc_nn)

# Establish a threshold at the optimal point on the ROC curve
threshold <- 0.5

# Use threshold to classify predictions
predicted_vax <- ifelse(predicted_prob > threshold, 1, 0)

# Create the confusion table
confusion_table <- table(Predicted = predicted_vax, Actual = data_test$flu_vax_30_days)
confusion_table

dput(confusion_table)

# Calculate the misclassification error
misclassification_error <- mean(predicted_vax != data_test$flu_vax_30_days)
print(paste("Misclassification Error:", misclassification_error))

```

## Graph ROC Curves
```{r}

# Generate png device
png("ROC-curve-comparison.png", width = 1200, height = 800, res = 150)

# Plot curve with false-positive on the x-axis
plot(1-roc_ols$specificities, roc_ols$sensitivities, col="#E64B35FF", pch=12, lty=1, cex=.1,
     xlab="False Positive",
     ylab="Sensitivity")
points(1-roc_logit$specificities, roc_logit$sensitivities, col="#DF8F44FF", pch=13, lty=2, cex = .1)
points(1-roc_relaxed_lasso_ols$specificities, roc_relaxed_lasso_ols$sensitivities, col="#008B45FF", pch=14, lty=3, cex = .1)
points(1-roc_relaxed_lasso_logit$specificities, roc_relaxed_lasso_logit$sensitivities, col="#7E6148FF", pch=15, lty=4, cex = .1)

points(1-roc_rf$specificities, roc_rf$sensitivities, col="#008280FF", pch=16, lty=5, cex = .1)
points(1-roc_nn$specificities, roc_nn$sensitivities, col="#3B4992FF", pch=17, lty=6, cex = .1)
legend("bottomright", legend = c("OLS", "Logit", "Relaxed Lasso - OLS", "Relaxed Lasso - Logit", "Random Forest", "Neural Net"),
       col = c("#E64B35FF", "#DF8F44FF", "#008B45FF", "#7E6148FF", "#008280FF", "#3B4992FF"), pch = c(12:17), lty = c(1:6))

dev.off()
```

# Validation

```{r}
# Obtain the predicted probabilities from the OLS model on validation set
predicted_prob <- predict(fit_ols, data_validate, type = "response")
predicted_prob <- ifelse(predicted_prob < 0, 0, predicted_prob)

# Plot ROC curve and calculate AUC 
roc_ols <- roc(data_validate$flu_vax_30_days, predicted_prob)
plot(roc_ols)

# Calculate the auc_value
auc_value <- auc(roc_ols)

# Establish a threshold at the optimal point on the ROC curve
threshold <- 0.5

# Use threshold to classify predictions  
predicted_vax <- ifelse(predicted_prob > threshold, 1, 0)

# Create the confusion table
confusion_table <- table(Predicted = predicted_vax, Actual = data_validate$flu_vax_30_days)
confusion_table

dput(confusion_table)

# Calculate the misclassification error
misclassification_error <- mean(predicted_vax != data_validate$flu_vax_30_days)
print(paste("Misclassification Error:", misclassification_error))

```